Peptide retention time prediction

### 2014-12-16 retention.tex
x

In this review we summarize the work done in the field of retention
time prediction, and describe a few of the successful applications of
such predictions. Sections~\ref{sec:rtpred} and \ref{sec:rtpredm}
describe the main approaches used to estimate retention time from
peptide sequences. Section~\ref{sec:app} discusses some of the most
important applications of retention time predictions.
Section~\ref{sec:irt} briefly introduces the use of peptide retention
standards.

@@

These methods involve keeping a database of previously observed
retention times for a set of peptides, and subsequently use these
values as a basis for retention time predictions. In particular, in
the past years the use of retention standards has become increasingly
widespread \cite{olegstd, irt}. Such standards typically consist of a
small number of peptides covering a wide span of hydrophobicities and
with well studied behaviors under LC-MS/MS. These peptides are often
analyzed together with the sample of interest, and by comparing their
observed and expected behaviors one can obtain valuable information
for a variety of applications.

@@

A limitation of SSRCalc is that the algorithm was optimized for
predicting retention times for certain chromatography setups. Thus,
SSRCalc's prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}.

@@

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM) experiments, are dependent on the predictability of a
peptides behavior. In such experiments the retention times and the
suitable transitions are needed as input for the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques techniques\cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiment
would lead to a more direct experimental design and hopefully to more
comprehensie coverage. Thus {\em de novo} retention time predictions
have an important role to fill.
%However, to do so we need to make the technique more accurate.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have put quite some
intest in so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines
(RBM)\cite{salakhutdinov2009deep} or deep neural networks. This
follows the observation that ideathe performance of predictors often
are enhanced if one can structure the prediction into multiple
layers. The drawback of such models are that they require large
training sets. Given the very large sets of peptides with retention
times that are assembled in various repositories, such as peptide
atlas and pride, such methods could probably be implemented for
retention time prediction as well.

>>>>

In this review we summarize the work done in the field of retention
time prediction, and describe a few of the successful applications of
such predictions. Sections~\ref{sec:rtpred} and \ref{sec:rtpredm}
describe the main approaches to retention time predictions from
peptide sequences. Section~\ref{sec:app} discusses some of the most
important applications of retention time predictions.
% Section~\ref{sec:irt} briefly introduces the use of peptide retention
% standards.

@@

These methods involve keeping a database of previously observed
retention times for a set of peptides, and subsequently use these
values as a basis for retention time predictions. In particular, in
the past years the use of retention standards has become increasingly
widespread \cite{olegstd, irt}. Such standards typically consist of a
small number of peptides covering a wide span of hydrophobicities and
favorable for detection by mass spectrometry. These peptides are often
analyzed together with the sample of interest, and by comparing their
observed and expected behaviors one can obtain valuable information
for a variety of applications.

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}.

@@

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM) experiments, are dependent on the predictability of a
peptides behavior. In such experiments the retention times and the
suitable transitions are needed as input for the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiment
would lead to a more direct experimental design and hopefully to more
comprehensie coverage. Thus {\em de novo} retention time predictions
have an important role to fill.
%However, to do so we need to make the technique more accurate.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have put quite some
intest in so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines
(RBM) \cite{salakhutdinov2009deep} or deep neural networks. This
follows the observation that ideathe performance of predictors often
are enhanced if one can structure the prediction into multiple
layers. The drawback of such models are that they require large
training sets. Given the very large sets of peptides with retention
times that are assembled in various repositories, such as peptide
atlas and pride, such methods could probably be implemented for
retention time prediction as well.
